---
title: "Week 11 Exercise"
author: "Alexandra Boville"
date: "2024-04-01"
output: html_document
---
Libraries Used
```{r}
library(car)
library(jtools)
library(tidyverse)
library(broom)
library(infer)
```

Loading in AVONET dataset
```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"

d <- read_csv(f, col_names =  TRUE)

d <- d %>% 
    select(c("Species1", "Family1", "Order1", "Beak.Length_Culmen", "Beak.Width", "Beak.Depth", "Tarsus.Length", "Wing.Length", "Tail.Length", "Mass", "Habitat", "Migration", "Trophic.Level", "Trophic.Niche", "Primary.Lifestyle", "Min.Latitude", "Max.Latitude", "Centroid.Latitude", "Range.Size"))

head(d)

summary(d)
#Species1, Family1, Order1, Habitat, Migration, Trophic.Level, Trophic.Niche, and Primary.Lifestyle are all categorical. Migration is scored as numeric and should be reclassified for analaysis
```
# Challenge 1 One-Factor ANOVA and Inference 

# Step 1: Making Boxplots of log(Mass) in relation to Trophic.Level and Migration
```{r}
#Organizing data for Boxplots
d <- d %>% 
  drop_na(Mass, Trophic.Level, Migration) %>% #removing rows with NA values in these columns
   mutate(Migration = as.factor(Migration)) #converting Migration variable from class chr to a factor

#Boxplot for log(Mass) in relation to Trophic.Level
ggplot(data = d %>% 
         drop_na(Trophic.Level), aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot() + geom_jitter()

#Boxplot for log(Mass) in relation to Migration
ggplot(data = d %>% 
         drop_na(Migration), aes(x = Migration, y = log(Mass))) +
  geom_boxplot() + geom_jitter()
```

# Step 2: Running Linear Models Analysis of Coefficients

```{r}
m <- lm(log(Mass) ~ Trophic.Level, data = d)
summary(m)

m1 <- lm(log(Mass) ~ Migration, data = d)
summary(m1)


```
For both linear models the p-value associated with the F statistic is less than 0.05 so log(Mass) is associated with Trophic.Level and Migration. The value of the F statistic is larger for the linear model looking at the relationship between log(Mass) and Migration and in general a larger F statistic value indicates there is more between group variance. To determine if this is significant when comparing both Trophic.Level and Migration you should run additional post-hoc analyses such as an ANOVA. 

Migration 1 is the reference level in this linear model and both Migration2 and Migration 3 are different from the reference level. 

```{r}
d_ref_relevel <- d %>%  
  mutate(Migration = relevel(Migration, ref = "2")) #switching ref = between 2 and 3 and running corresponding lm() below to see how this alters coefficients

m1_relevel <- lm(log(Mass) ~ Migration, data = d_ref_relevel)
summary(m1_relevel)

```
Changing the reference level to Migration2 from Migration1 results in the negative estimate for Migration1 suggesting that Migration is negatively correlated with log(Mass). Changing the reference level from Migration1 to Migration3 results in estimates for both remaining coefficients as negative values.


# Step 3: Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which Migration categories differ “significantly” from one another
```{r}
m1aov <- aov(log(Mass) ~ Migration, data = d)
summary(m1aov)

(posthoc <- TukeyHSD(m1aov, which = "Migration", conf.level = 0.95)) #produces adjusted p value for each posthoc comparison

#Migration2 and Migration1 differ most significantly from one another
```

# Permutation approach to inference to generate a null distribution of F statistic values for the model of log(Mass) in relation to Trophic.Level
```{r}
original.F <- aov(log(Mass) ~ Trophic.Level, data = d) %>% 
  tidy() %>% 
  filter(term == "Trophic.Level")
original.F #results for F stat and p value for omnibus F test


d <- d %>% 
  mutate(logMass = log(Mass))

permuted.F <- d %>% 
  specify(logMass ~ Trophic.Level) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "F")

str(permuted.F)
 
visualize(permuted.F) +
  shade_p_value(obs_stat = original.F$statistic, direction = "greater") #value of f statistic way way far to right
#p value ver y very tiny

permuted.F
```

# Challenge 2: Data Wrangling, One- and Two-Factor ANOVA

# Step 1: Creating Variables relative beak length and relative tarsus length

```{r}
d$relBL <- resid(lm(formula = log(Beak.Length_Culmen) ~ log(Mass), data = d))
d$relTL <- resid(lm(formula = log(Tarsus.Length) ~ log(Mass), data = d))
```

# Step 2: Boxplot of relative tarsus length in relation to Primary.Lifestyle and relative beak length in relation to Trophic.Niche

```{r}
ggplot(data = d %>% 
         filter(!is.na(Primary.Lifestyle)),
       aes(x = Primary.Lifestyle, y = relTL)) +
  geom_boxplot() 

ggplot(data = d %>% 
         filter(!is.na(Trophic.Niche)),
       aes(x = Trophic.Niche, y = relBL)) +
  geom_boxplot() 
```

# Step 3: ANOVA analysis to look at association between Range.Size and Migration
```{r}
#already dropped NA values for migration and converted to as.factor
m1aov <- aov(Range.Size ~ Migration, data = d)
summary(m1aov)

TukeyHSD(m1aov)


d_ref_relevel <- d %>%  
  mutate(Migration = relevel(Migration, ref = "3")) #switching ref = between 2 and 3 and running corresponding lm() below to see how this alters coefficients

m1_relevel <- lm(Range.Size ~ Migration, data = d_ref_relevel)
summary(m1_relevel)


TukeyHSD(m1aov)

```
Range size is significantly dependent on Migration. All migration factors are different from the reference level. The reference level is Migration1 (sedentary lifestyle) but all Migration levels are significant.  

# Step 4: Looking at Only Passeriformes Data

```{r}
#Reducing Data to only include members of Order Passeriformed
passeriformes <- d %>% 
  filter(Order1 == "Passeriformes")

head(passeriformes)
```

```{r}
#Running one-factor ANOVA
aov1 <- aov(relBL ~ Trophic.Level, data = passeriformes)
aov2 <- aov(relBL ~ Primary.Lifestyle, data = passeriformes)
summary(aov1)
summary(aov2)

#boxplots of response variable by each predictor and by the combination of predictors
ggplot(passeriformes %>% 
         filter(!is.na(Trophic.Level)), aes(x = Trophic.Level, y =relBL)) + 
        geom_boxplot()

ggplot(passeriformes %>% 
         filter(!is.na(Trophic.Level)), aes(x = Primary.Lifestyle, y =relBL)) + 
        geom_boxplot()
```
Both Trophic.Level and Primary.Lifestyle are good predictors of relatuve beak length. Both have siginficant P values and high F-values. 

# Step 5: Running Two-factor ANOVA

```{r}

m_passeriformes <- aov(relBL ~ Trophic.Level + Primary.Lifestyle, data = passeriformes) #looking at relative association between Tropic.Level and Primary.Lifestyle
summary(m_passeriformes)

```
Tropic.Level has a larger effect on relative beak length since it has a larger F value. 

#Step 6: Two-factor ANOVA with interaction of predictor variables
```{r}
m_t_and_p_interact <- aov(relBL ~ Trophic.Level + Primary.Lifestyle + Trophic.Level:Primary.Lifestyle, data = passeriformes) # : indicated interaction
summary(m_t_and_p_interact)
```
The interaction between Trophic.Level and Primary.Lifestyle do not predict relative beak length more than each predictor variable does independently. 


```{r}
interaction.plot(
  x.factor = passeriformes$Trophic.Level,
  xlab = "Trophic Level",
  trace.factor = passeriformes$Primary.Lifestyle,
  trace.label = "Primary Lifestyle",
  response = passeriformes$relBL,
  ylab = "Relative Beak Length"
)
```










